Synthesizing Quotient Lenses

### 2018-02-21 papers/QuotientLenses/main.tex
section 2 edits checkpoint

\section{Quotient Regular Expressions and Lenses by Example}
\label{sec:example}

@@

Consider a bidirectional transformation that converts
\bibtex{} citation records such as
\begin{verbatim}
@Book {conway,
Author = "J. H. Conway",
Title = {Regular Algebra and Finite Machines}
}
\end{verbatim}
\noindent
to an equivalent EndNote format like the following
\begin{verbatim}
%0 Book
%T Regular Algebra and Finite Machines
%A J. H. Conway
%F conway
\end{verbatim}
\noindent
and vice versa.
%
Transformations from one such representation to the other will preserve
most of the information content of the data:  the title, the author,
and the key, for instance.  However, intuitively,
some of the information is inessential and may be discarded---indeed, this
inessential information will likely appear so trivial that a casual
reader may not even realize it is there to begin with.  For instance,
there is no need to preserve the amount of whitespace between words, at the ends of lines, or
between fields.  In addition, the order of author and title fields is irrelevant.
There may also be additional equivalent ways of writing the same
name:  ``Conway, J. H.'' and ``J. H. Conway'' should probably be treated
the same.

@@

When two instances of the same format differ only terms of such inessential
information, it is often useful to treat them as identical.
For example, the following two
\bibtex{} citations represent the same logical object and as a result, should
be translated to the same EndNote record.
\begin{verbatim}
@Book {conway,
Author = "J. H. Conway",
Title = {Regular Algebra and Finite Machines},
}

@Book{     conway,
Title = {Regular Algebra
and Finite Machines},
Author = "Conway, J. H.",     }
\end{verbatim}
\noindent EndNote records should also have a similar equivalence relation
defined on them.

While it is possible to express the equivalence on the \bibtex{} data above in
Boomerang, actually doing so is a challenge.  Three of the inessential details
we wish to ignore---whitespace quotienting
inside fields, whitespace quotienting in between fields, and permutation
quotenting of fields---nest within each other in a way that makes them difficult
to express with the existing Boomerang combinators. Furthermore, if the
programmer wishes to specify the type of the lens, then she can only specify
data formats on each side of the lens with a regular expression that does not
keep track of the equivalence relation defined on the language of the regular
expression. Consequently, the programmer has to write a regular expression as
well as a canonizer that expresses the equivalence relation on that regular
expression, even though a regular expression and the canonizer defined on it
typically share much of the syntactic structure. Having to write both of these
expressions makes programming with quotient lenses verbose and time consuming.

To address these issues, we propose {\em Quotient Regular Expressions}, or QREs,
which enable the programmer to define a regular expression $R$, an equivalence
relation $\sim$ on the language described by $R$ as well as a canonical set
of representatives of $\sim$, all in one syntactic term. For example, with QREs
we can specify \bibtex{} records and their canonical elements, as shown in
Figure~\ref{fig:example-qre}.

>>>>

\section{Quotients by Example}
\label{sec:example}

@@

\emph{Lenses} define computations that can be read in two directions, forwards and
backwards.  The simplest lenses define bijections between source and
data, but we are interested in \emph{quotient lenses}, which define
bijections modulo some equivalence relation.  The remains of this section
describe and motivate a new language and system for defining such lenses.

To begin, consider a bidirectional transformation that converts
\bibtex{} citation records such as
\begin{verbatim}
@Book {conway,
Author = "J. H. Conway",
Title = {Regular Algebra and Finite Machines}
}
\end{verbatim}
\noindent
into an equivalent EndNote format like the following
\begin{verbatim}
%0 Book
%T Regular Algebra and Finite Machines
%A J. H. Conway
%F conway
\end{verbatim}
\noindent
and vice versa.
%
Notice that transformations from one format to the other preserve
most of the information present in the data:  the title, the author,
and the key, for instance.  However, some information in the syntax
may be lost.  For instance, users are typically uninterested in preserving
the amount of the whitespace between words or
between fields.  The order of author and title fields is also likely
irrelevant and there may be equivalent ways of writing the same
name:  ``Conway, J. H.'' vs ``J. H. Conway.''

@@

When two instances of the same format differ only terms of such inessential
information, it is often useful to treat them as identical and to map them
into canonical representations.
Continuing our example, the following two
\bibtex{} citations represent the same logical object and as a result, should
be translated to the same EndNote record.
\begin{verbatim}
@Book {conway,
Author = "J. H. Conway",
Title = {Regular Algebra and Finite Machines},
}

@Book{     conway,
Title = {Regular Algebra
and Finite Machines},
Author = "Conway, J. H.",     }
\end{verbatim}

One natural way to define lenses that will synchronize data in these formats
is to provide the following items:

\begin{enumerate}
\item A pair of descriptions of the external source and target formats.
Such descriptions may be given by regular expressions, for instance.
They serve as documentation to describe the domains over which the lens
is to operate.
\item A pair of \emph{canonizers} that map external source (and target)
objects into internal, canonical forms.  Such canonizers will take
care of stripping away irrelevant information and otherwise normalizing
differences between equivalent representations of the same logical data.
When given an internal representation, such canonizers must also operate
in the reverse direction, \emph{choosing} an element in the external format
to represent the canonical internal element.
\item A pair of descriptions of the internal source and
target formats, again, to document the structure of the transformation.
\item A bijective lens to map back and forth between canonical elements
of the internal source and target formats.
\end{enumerate}

\noindent
Indeed, Boomerang's quotient lenses are structured this way~\ref{boomerang}.
Of course, some of these elements are optional.  For instance, it is
not necessary to document internal and external formats.  However, we do
view it as good engineering practice to do so, as such documentation makes
lenses easier to read, understand, use and maintain.  In this case, there is
significant redundancy here:  External and internal formats
typically have some overlap; external formats and canonizers have related
structure; and the structure of bijective lens computations is largely driven
by the structure of the internal formats.  Hence, overall, a more compact
notation would simplify description and legibility of such computations.
In addition,
the lens and canonizer languages themselves can be somewhat fiddly and
difficult to
master.  For instance, to quotient out whitespace inside, and in between,
fields while also ignoring field order one must nest canonizing combinators
within one another in a somewhat tricky and fiddly way.

%% %Foster's Boomerang~\ref{boomerang} language presents one design
%% While it is possible to express the equivalence on the \bibtex{} data above in
%% Boomerang, actually doing so is a challenge.  Three of the inessential details
%% we wish to ignore---whitespace quotienting
%% inside fields, whitespace quotienting in between fields, and permutation
%% quotenting of fields---nest within each other in a way that makes them difficult
%% to express with the existing Boomerang combinators. Furthermore, if the
%% programmer wishes to specify the type of the lens, then she can only specify
%% data formats on each side of the lens with a regular expression that does not
%% keep track of the equivalence relation defined on the language of the regular
%% expression. Consequently, the programmer has to write a regular expression as
%% well as a canonizer that expresses the equivalence relation on that regular
%% expression, even though a regular expression and the canonizer defined on it
%% typically share much of the syntactic structure. Having to write both of these
%% expressions makes programming with quotient lenses verbose and time consuming.

We propose to do away with much of this overhead and the burden of a
fiddly programming process by defining a single new compact notation,
the \emph{Quotient Regular Expression}, which simultaneously describes
external format, equivalence relation and canonical forms, and then to
synthesize lenses that map between these objects automatically.

\subsection{Quotient Regular Expressions}

A {\em Quotient Regular Expression} (QRE) defines an external regular
expression $R$,
an equivalence relation $\sim$ on the language described by $R$, and a canonical set
of representatives of $\sim$, all in one syntactic term.
Figure~\ref{fig:example-qre} illustrates how to 
specify \bibtex{} records and their canonical elements using QREs.
